---
title: <center> My Professional Website
---

<center>

### **About**

With a strong background in IT program and project management, I excel at delivering transformative enterprise solutions and driving digital innovation across diverse industries. I have successfully led high-impact initiatives, overseeing P&L, managing cross-functional programs, and building high-performing teams to deliver scalable, results-driven solutions. My expertise includes strategic planning, end-to-end project delivery, and process optimization, ensuring complex programs are executed efficiently and achieve measurable value. By focusing on innovation, collaboration, and operational excellence, I drive impactful outcomes that align with organizational goals and create lasting success.

[Learn more about my professional background](https://www.linkedin.com/in/kyleshaddox/ "Open LinkedIn Profile")

</center>

<center>

### I am also a student at Baruch University

</center>

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```

<Center>

```{r}
#| include: false
1+1
```

------------------------------------------------------------------------

Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`

</Center>
